---
title: "Does the Test Actually Classify People Correctly?"
subtitle: "PSYC 7304/8304 · Diagnostic Accuracy and ROC Analysis"
author: "Ryan Farmer, PhD"
format:
  revealjs:
    theme: [default, week12_custom.scss]
    slide-number: true
    progress: true
    history: false
    transition: fade
    transition-speed: fast
    background-transition: fade
    code-fold: false
    toc: false
    center: false
    width: 1200
    height: 700
    margin: 0.06
    smaller: false
    highlight-style: github
    embed-resources: true
    preview-links: false
    chalkboard: false
    mouse-wheel: false
    controls: true
    controls-layout: bottom-right
    fragments: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(pROC)
library(knitr)
library(kableExtra)
library(scales)
library(gridExtra)

# ── nomogrammer (local copy, Chekroud 2016, MIT License) ─────────────────────
nomogrammer <- function(Prevalence, Plr = NULL, Nlr = NULL,
                        Sens = NULL, Spec = NULL,
                        Detail = FALSE, NullLine = FALSE,
                        LabelSize = (14 / 5), Verbose = FALSE) {
  if (!is.null(Sens) & !is.null(Spec)) {
    Plr <- Sens / (1 - Spec); Nlr <- (1 - Sens) / Spec
  }
  PreTestOdds     <- Prevalence / (1 - Prevalence)
  PostTestOddsPos <- PreTestOdds * Plr
  PostTestOddsNeg <- PreTestOdds * Nlr
  PostTestProbPos <- PostTestOddsPos / (1 + PostTestOddsPos)
  PostTestProbNeg <- PostTestOddsNeg / (1 + PostTestOddsNeg)
  probs <- c(0.001,0.002,0.003,0.005,0.007,0.01,0.02,0.03,0.05,0.07,
             0.10,0.15,0.20,0.30,0.40,0.50,0.60,0.70,0.80,0.85,
             0.90,0.93,0.95,0.97,0.98,0.99,0.993,0.995,0.997,0.999)
  lodds       <- function(p) log(p / (1 - p))
  axis_lo     <- lodds(probs)
  axis_labels <- percent(probs, accuracy = 0.1)
  axis_labels <- gsub("^0\\.", ".", axis_labels)
  lr_vals  <- c(0.001,0.002,0.003,0.005,0.007,0.01,0.02,0.03,0.05,0.07,
                0.1,0.2,0.3,0.5,0.7,1,2,3,5,7,10,20,30,50,70,
                100,200,300,500,1000)
  lr_scale <- log(lr_vals); lr_labels <- as.character(lr_vals)
  x_pre <- 0; x_lr <- 1; x_post <- 2
  df_pre  <- data.frame(x = x_pre,  y = axis_lo,  lab = axis_labels)
  df_lr   <- data.frame(x = x_lr,   y = lr_scale, lab = lr_labels)
  df_post <- data.frame(x = x_post, y = axis_lo,  lab = axis_labels)
  pre_y      <- lodds(Prevalence)
  post_y_pos <- lodds(PostTestProbPos)
  post_y_neg <- lodds(PostTestProbNeg)
  lr_y_pos   <- log(Plr); lr_y_neg <- log(Nlr)
  p <- ggplot() +
    geom_point(data=df_pre,  aes(x=x,y=y),size=0.8,color="#555") +
    geom_text(data=df_pre,   aes(x=x-0.06,y=y,label=lab),
              size=LabelSize,hjust=1,color="#333") +
    geom_point(data=df_lr,   aes(x=x,y=y),size=0.8,color="#555") +
    geom_text(data=df_lr,    aes(x=x+0.06,y=y,label=lab),
              size=LabelSize,hjust=0,color="#333") +
    geom_point(data=df_post, aes(x=x,y=y),size=0.8,color="#555") +
    geom_text(data=df_post,  aes(x=x+0.06,y=y,label=lab),
              size=LabelSize,hjust=0,color="#333") +
    geom_line(data=df_pre,  aes(x=x,y=y),color="#555",linewidth=0.4) +
    geom_line(data=df_lr,   aes(x=x,y=y),color="#555",linewidth=0.4) +
    geom_line(data=df_post, aes(x=x,y=y),color="#555",linewidth=0.4) +
    {if(NullLine) geom_segment(aes(x=x_pre,xend=x_post,y=pre_y,yend=pre_y),
                               color="#bbb",linewidth=0.7,linetype="dashed")} +
    geom_segment(aes(x=x_pre,xend=x_post,y=pre_y,yend=post_y_pos),
                 color="#2E4057",linewidth=1.2) +
    geom_point(aes(x=x_pre, y=pre_y),     color="#2E4057",size=3.5) +
    geom_point(aes(x=x_lr,  y=lr_y_pos),  color="#2E4057",size=3.5) +
    geom_point(aes(x=x_post,y=post_y_pos),color="#2E4057",size=3.5,shape=17) +
    geom_segment(aes(x=x_pre,xend=x_post,y=pre_y,yend=post_y_neg),
                 color="#E07B39",linewidth=1.2) +
    geom_point(aes(x=x_lr,  y=lr_y_neg),  color="#E07B39",size=3.5) +
    geom_point(aes(x=x_post,y=post_y_neg),color="#E07B39",size=3.5,shape=17) +
    {if(Detail) list(
      # Pre-test label offset left to avoid axis collision
      annotate("text",x=x_pre-0.20,y=pre_y+0.45,
               label=percent(Prevalence,accuracy=0.1),
               color="#2E4057",size=LabelSize+0.5,fontface="bold",hjust=1),
      # LR labels offset right, with vertical nudge to avoid axis ticks
      annotate("text",x=x_lr+0.18,y=lr_y_pos+0.55,
               label=paste0("LR+ = ",round(Plr,2)),
               color="#2E4057",size=LabelSize+0.3,hjust=0),
      annotate("text",x=x_lr+0.18,y=lr_y_neg-0.55,
               label=paste0("LR- = ",round(Nlr,2)),
               color="#E07B39",size=LabelSize+0.3,hjust=0),
      # Post-test labels: nudge up/down away from axis ticks
      annotate("text",x=x_post+0.18,y=post_y_pos+0.55,
               label=percent(PostTestProbPos,accuracy=0.1),
               color="#2E4057",size=LabelSize+0.5,fontface="bold",hjust=0),
      annotate("text",x=x_post+0.18,y=post_y_neg-0.55,
               label=percent(PostTestProbNeg,accuracy=0.1),
               color="#E07B39",size=LabelSize+0.5,fontface="bold",hjust=0)
    )} +
    annotate("text",x=x_pre, y=max(axis_lo)+0.9,label="Pre-test\nprobability",
             size=LabelSize+0.6,fontface="bold",color="#333",hjust=0.5) +
    annotate("text",x=x_lr,  y=max(axis_lo)+0.9,label="Likelihood\nratio",
             size=LabelSize+0.6,fontface="bold",color="#333",hjust=0.5) +
    annotate("text",x=x_post,y=max(axis_lo)+0.9,label="Post-test\nprobability",
             size=LabelSize+0.6,fontface="bold",color="#333",hjust=0.5) +
    scale_x_continuous(limits=c(-0.65,2.85),expand=c(0,0)) +
    theme_void() + theme(plot.margin=margin(12,8,12,8))
  return(p)
}

# ── Simulated data ────────────────────────────────────────────────────────────
set.seed(42)

sim_covid <- function(n=1000, prevalence, sens=0.86, spec=0.99, seed_offset=0) {
  set.seed(42 + seed_offset)
  n_pos <- rbinom(1, n, prevalence); n_neg <- n - n_pos
  tp <- rbinom(1, n_pos, sens);      fn <- n_pos - tp
  fp <- rbinom(1, n_neg, 1-spec);    tn <- n_neg - fp
  tibble(true_status = c(rep("Positive",n_pos), rep("Negative",n_neg)),
         test_result = c(rep("Positive",tp), rep("Negative",fn),
                         rep("Positive",fp), rep("Negative",tn)))
}

covid_mid <- sim_covid(n=1000, prevalence=0.30, seed_offset=2)

n_phq <- 500; prev_phq <- 0.25
n_dep <- round(n_phq * prev_phq); n_nodep <- n_phq - n_dep
set.seed(42)
phq_dep   <- pmax(0, pmin(27, round(rnorm(n_dep,   15.0, 4.0))))
phq_nodep <- pmax(0, pmin(27, round(rnorm(n_nodep,  5.5, 3.5))))

dat_phq <- tibble(
  score       = c(phq_dep, phq_nodep),
  true_status = c(rep("Depressed", n_dep), rep("Not depressed", n_nodep))
) |> mutate(test_pos = score >= 10)

compute_dx_stats <- function(tp, fp, fn, tn, prevalence=NULL) {
  sens <- tp/(tp+fn); spec <- tn/(tn+fp)
  prev <- if(!is.null(prevalence)) prevalence else (tp+fn)/(tp+fp+fn+tn)
  list(sens=sens, spec=spec,
       ppv=(sens*prev)/(sens*prev+(1-spec)*(1-prev)),
       npv=(spec*(1-prev))/((1-sens)*prev+spec*(1-prev)),
       lrp=sens/(1-spec), lrn=(1-sens)/spec)
}

phq_tp <- sum(dat_phq$true_status=="Depressed"    &  dat_phq$test_pos)
phq_fn <- sum(dat_phq$true_status=="Depressed"    & !dat_phq$test_pos)
phq_fp <- sum(dat_phq$true_status=="Not depressed" &  dat_phq$test_pos)
phq_tn <- sum(dat_phq$true_status=="Not depressed" & !dat_phq$test_pos)
phq_st <- compute_dx_stats(phq_tp, phq_fp, phq_fn, phq_tn, prevalence=0.25)

cov_tab <- covid_mid |>
  summarise(tp=sum(true_status=="Positive"&test_result=="Positive"),
            fn=sum(true_status=="Positive"&test_result=="Negative"),
            fp=sum(true_status=="Negative"&test_result=="Positive"),
            tn=sum(true_status=="Negative"&test_result=="Negative"))
cov_st <- compute_dx_stats(cov_tab$tp,cov_tab$fp,cov_tab$fn,cov_tab$tn,prevalence=0.30)

roc_phq <- roc(response=ifelse(dat_phq$true_status=="Depressed",1,0),
               predictor=dat_phq$score, direction="<", quiet=TRUE)
auc_val <- auc(roc_phq)
ci_val  <- ci.auc(roc_phq)
```

## The question we haven't asked yet {background-color="#2E4057"}

::: {style="font-size:1.05em; margin-top:0.8em; color:#ffffff;"}
We have asked: does the test measure what it claims to measure?

We have asked: does it measure the same thing for everyone?

**We have not yet asked:** when a clinician uses a test score to make a decision, how often does the decision turn out to be correct?
:::

::: {style="background:#1a2d40; border-left:4px solid #7fa8c9; padding:0.6em 0.9em; margin-top:1.1em; font-size:0.88em; color:#d0dde8;"}
This is the question diagnostic accuracy methods answer. It is distinct from reliability, factor structure, and correlational validity — and it is the one practitioners most need answered.
:::

::: notes
Opening framing. The previous chapter was bias/DIF. This chapter shifts from "does the test measure fairly" to "does the test classify correctly." Set up the gap between what test manuals report and what clinicians need.
:::

## Two instruments, same manual statistics, very different classifiers

:::: {.columns}
::: {.column width="50%"}
**Instrument A**

- Internal consistency: α = .92
- Criterion correlation: *r* = .70
- Sensitivity at cutoff: **.80**
- False positive rate: **.05**
:::

::: {.column width="50%"}
**Instrument B**

- Internal consistency: α = .91
- Criterion correlation: *r* = .69
- Sensitivity at cutoff: **.60**
- False positive rate: **.15**
:::
::::

::: {style="margin-top:1em; font-size:0.9em; color:#555;"}
On the properties typically reported in test manuals, these instruments are nearly identical.
On classification performance, Instrument B is substantially inferior.
:::

::: notes
Motivating example. Students should recognize that the statistics they have spent weeks learning are necessary but not sufficient for evaluating clinical utility.
:::

# The Classification Problem {background-color="#2E4057"}

---

## The confusion matrix

```{r}
tribble(
  ~` `,                ~`Condition present`,  ~`Condition absent`,
  "Test positive",     "True Positive (TP)",  "False Positive (FP)",
  "Test negative",     "False Negative (FN)", "True Negative (TN)"
) |>
  kable() |>
  kable_styling(bootstrap_options=c("striped","hover"),
                full_width=FALSE, font_size=24) |>
  row_spec(0, background="#2E4057", color="white") |>
  column_spec(1, bold=TRUE, background="#f0f4f8")
```

::: {style="margin-top:0.9em; font-size:0.88em;"}
Every classification decision — a COVID test, a depression screen, a reading probe, a neuropsychological battery — produces exactly one of these four outcomes. The structure is identical across domains.
:::

::: notes
Establish the universal structure. Students have encountered 2x2 tables in research methods; connect to that prior knowledge.
:::

---

## COVID rapid antigen test at 30% prevalence

```{r}
covid_mid |>
  mutate(true_status=factor(true_status,levels=c("Positive","Negative")),
         test_result=factor(test_result,levels=c("Positive","Negative"))) |>
  count(test_result, true_status) |>
  pivot_wider(names_from=true_status, values_from=n, values_fill=0) |>
  kable(col.names=c("Test result","Condition present","Condition absent")) |>
  kable_styling(bootstrap_options=c("striped","hover"),
                full_width=FALSE, font_size=24) |>
  row_spec(0, background="#2E4057", color="white") |>
  column_spec(1, bold=TRUE)
```

::: {style="margin-top:0.8em; font-size:0.9em;"}
Overall accuracy: **`r round((filter(covid_mid, true_status==test_result)|>nrow())/nrow(covid_mid)*100,1)`%** — sounds good. But most correct classifications are negative cases correctly cleared.
:::

::: notes
Walk through the table. Ask: what does a positive result mean for a specific patient? This sets up the base rate section.
:::

---

## Why overall accuracy misleads

::: {style="font-size:0.95em; margin-bottom:0.7em;"}
A test that calls **everyone negative** in a 5% prevalence setting achieves **95% accuracy** — with zero sensitivity.

Overall accuracy is dominated by the most common outcome. In screening contexts, that means it is dominated by true negatives, not by the cases you are trying to detect.
:::

::: {.callout-important}
Overall accuracy cannot tell you how well a test performs on the group you most care about detecting. It is almost always the wrong summary statistic for a screening or diagnostic instrument.
:::

::: notes
This is the most common misconception in applied assessment. Students who have read test manuals will have seen "hit rate" reported prominently. This slide is the corrective.
:::

---

## No cutoff is perfect

```{r fig-cutoffs-slides, fig.height=4.6, fig.width=11.0, out.width="100%"}
cutoff_labels <- c("Lenient (7): high Se, lower Sp",
                   "Standard (10): balanced",
                   "Strict (13): high Sp, lower Se")
cutoff_vals   <- c(7, 10, 13)

score_seq  <- seq(0, 27, length.out=400)
dens_dep   <- dnorm(score_seq, 15.0, 4.0)
dens_nodep <- dnorm(score_seq,  5.5, 3.5)
max_dens   <- max(c(dens_dep, dens_nodep))

dens_df <- bind_rows(
  tibble(score=score_seq, density=dens_dep,   group="Depressed"),
  tibble(score=score_seq, density=dens_nodep, group="Not depressed")
)

plots_list <- lapply(seq_along(cutoff_vals), function(i) {
  co <- cutoff_vals[i]
  fp_shade <- tibble(score   = score_seq[score_seq >= co],
                     density = dnorm(score_seq[score_seq >= co], 5.5, 3.5))
  fn_shade <- tibble(score   = score_seq[score_seq <  co],
                     density = dnorm(score_seq[score_seq <  co], 15.0, 4.0))

  # Robust label positions that work for all three cutoffs
  tn_x  <- max(0.5, co * 0.35)
  tn_y  <- dnorm(tn_x, 5.5, 3.5) * 0.45
  fn_x  <- max(0.5, co * 0.55)
  fn_y  <- dnorm(fn_x, 15.0, 4.0) * 0.40
  tp_x  <- min(26.5, co + (27 - co) * 0.55)
  tp_y  <- dnorm(tp_x, 15.0, 4.0) * 0.50
  fp_x  <- min(26.5, co + (27 - co) * 0.25)
  fp_y  <- dnorm(fp_x, 5.5, 3.5) * 1.8

  ggplot() +
    geom_ribbon(data=fp_shade, aes(x=score,ymin=0,ymax=density),
                fill="#E07B39", alpha=0.40) +
    geom_ribbon(data=fn_shade, aes(x=score,ymin=0,ymax=density),
                fill="#2E4057", alpha=0.40) +
    geom_line(data=dens_df, aes(x=score,y=density,color=group), linewidth=1.0) +
    geom_vline(xintercept=co, linetype="dashed", color="#333", linewidth=0.9) +
    annotate("text", x=tn_x, y=tn_y, label="TN",
             color="#E07B39", size=3.8, fontface="bold") +
    annotate("text", x=fn_x, y=fn_y, label="FN",
             color="#2E4057", size=3.8, fontface="bold") +
    annotate("text", x=tp_x, y=tp_y, label="TP",
             color="#2E4057", size=3.8, fontface="bold") +
    annotate("text", x=fp_x, y=fp_y, label="FP",
             color="#E07B39", size=3.8, fontface="bold") +
    scale_color_manual(values=c("Depressed"="#2E4057","Not depressed"="#E07B39")) +
    scale_x_continuous(limits=c(0,27), breaks=seq(0,25,5)) +
    labs(x="Score", y="", subtitle=cutoff_labels[i]) +
    theme_minimal(base_size=11) +
    theme(legend.position  = if(i==1) "top" else "none",
          legend.title     = element_blank(),
          panel.grid.minor = element_blank(),
          plot.subtitle    = element_text(size=9.5, color="#444"))
})

grid.arrange(plots_list[[1]], plots_list[[2]], plots_list[[3]], ncol=3)
```

::: notes
TP/FP/FN/TN labels are positioned relative to each cutoff value so they don't overlap the curves. Moving the cutoff trades one error type for the other; this motivates ROC analysis.
:::


# Sensitivity and Specificity {background-color="#2E4057"}


## Two conditional probabilities

::: {style="font-size:0.92em; margin-bottom:0.5em;"}
**Sensitivity** answers: *given that the person has the condition*, what is the probability of a positive result?

$$\text{Sensitivity} = P(\text{Test}+ \mid \text{Condition}+) = \frac{TP}{TP + FN}$$

**Specificity** answers: *given that the person does not have the condition*, what is the probability of a negative result?

$$\text{Specificity} = P(\text{Test}- \mid \text{Condition}-) = \frac{TN}{TN + FP}$$
:::

::: {.callout-note}
Both are conditioned on **known true status**. They describe test performance looking forward from an already-established ground truth. They do **not** tell a clinician what a result means for a specific patient.
:::

::: notes
Emphasize the conditional probability framing. The "looking forward from ground truth" language helps. Foreshadow that PPV/NPV reverse the conditioning direction.
:::

---

## SnNOut / SpPIn

:::: {.columns}
::: {.column width="50%"}
**Sensitive test, Negative result**

**Rules Out**

If a test is highly sensitive, nearly all people with the condition test positive. A negative result therefore almost certainly means the condition is absent.

*COVID antigen: Se = `r percent(cov_st$sens, accuracy=0.1)`*

A negative result does not fully rule out infection.
:::

::: {.column width="50%"}
**Specific test, Positive result**

**Rules In**

If a test is highly specific, nearly all positive results come from people who truly have the condition. A positive result is strong evidence of the condition.

*COVID antigen: Sp = `r percent(cov_st$spec, accuracy=0.1)`*

A positive result strongly rules in infection.
:::
::::

::: {style="font-size:0.78em; color:#666; margin-top:0.7em;"}
Mnemonic from Sackett et al. (1991). Screening calls for high sensitivity; confirmation calls for high specificity.
:::

::: notes
The COVID example is intuitive. Students know that negative rapid tests during surges needed to be repeated — that is SnNOut failing because sensitivity was imperfect.
:::

---

## Test performance vs. decision performance

::: {style="font-size:0.88em; margin-bottom:0.6em;"}
Sensitivity and specificity describe **test performance** — how the instrument behaves when true status is already known.

Clinicians work in the opposite direction: they have a result and need to know what it implies about a patient they have not yet diagnosed.

**That is a different conditional probability. It requires different statistics.**
:::

```{r}
tribble(
  ~Question,                                      ~`Conditioned on`,    ~Statistic,
  "Of sick people, how many does the test catch?", "Known true status",  "Sensitivity",
  "Of healthy people, how many does it clear?",    "Known true status",  "Specificity",
  "This person tested positive — are they sick?",  "Known test result",  "PPV",
  "This person tested negative — are they well?",  "Known test result",  "NPV"
) |>
  kable() |>
  kable_styling(bootstrap_options=c("striped","hover"),
                full_width=TRUE, font_size=19) |>
  row_spec(0, background="#2E4057", color="white") |>
  row_spec(3:4, background="#FFF2CC")
```

::: notes
This table is the conceptual pivot. Top two rows = what we covered. Bottom two (highlighted) = what comes next. The distinction between these conditional probability spaces is the most important single idea in the chapter.
:::

# Predictive Values and Base Rates {background-color="#2E4057"}


## PPV and NPV: conditioned on the test result

::: {style="font-size:0.90em; margin-bottom:0.5em;"}
**Positive predictive value (PPV):** given a positive result, what is the probability the person has the condition?

$$PPV = P(\text{Condition}+ \mid \text{Test}+) = \frac{TP}{TP + FP}$$

**Negative predictive value (NPV):** given a negative result, what is the probability the person does not have the condition?

$$NPV = P(\text{Condition}- \mid \text{Test}-) = \frac{TN}{TN + FN}$$
:::

::: {.callout-important}
PPV and NPV are **not fixed properties of the test**. They depend on sensitivity, specificity, *and* the base rate of the condition in the population being tested.
:::

::: notes
The callout is the key teaching point. Students often assume PPV and NPV are properties of the instrument. The next slide demonstrates directly why they are not.
:::

---

## The base rate problem

```{r}
tibble(
  `Prevalence`  = c("5% — community screening",
                    "30% — household contact",
                    "60% — symptomatic cluster"),
  `Sensitivity` = "86%",
  `Specificity` = "99%",
  PPV = c(
    percent((0.86*.05)/(0.86*.05+0.01*.95), accuracy=0.1),
    percent((0.86*.30)/(0.86*.30+0.01*.70), accuracy=0.1),
    percent((0.86*.60)/(0.86*.60+0.01*.40), accuracy=0.1)
  ),
  NPV = c(
    percent((0.99*.95)/(0.14*.05+0.99*.95), accuracy=0.1),
    percent((0.99*.70)/(0.14*.30+0.99*.70), accuracy=0.1),
    percent((0.99*.40)/(0.14*.60+0.99*.40), accuracy=0.1)
  )
) |>
  kable() |>
  kable_styling(bootstrap_options=c("striped","hover"),
                full_width=FALSE, font_size=21) |>
  row_spec(0, background="#2E4057", color="white") |>
  row_spec(1, background="#FCE4D6") |>
  column_spec(1, bold=TRUE)
```

::: {style="font-size:0.88em; margin-top:0.6em;"}
The test does not change. Sensitivity and specificity are identical in all three rows. Only the population changes — and that alone drives PPV from 81% to 99%.
:::

::: notes
Work through the arithmetic for the red row: at 5% prevalence, 50 of 1000 are truly positive; the test finds about 43 (86% sensitivity). But it also flags about 10 of the 950 negatives (1% false positive rate). So PPV = 43/(43+10) = 81%.
:::

---

## Why PPV rises sharply but NPV declines slowly

::: {style="font-size:0.92em; margin-bottom:0.6em;"}
Both follow from the structure of Bayes' theorem.

**PPV** has true positives in its numerator and all positives (true + false) in its denominator. When prevalence is low, most positives are false — the denominator swamps the numerator. **PPV falls sharply.**

**NPV** draws its numerator from the large non-diseased majority, which contributes stably regardless of prevalence. **NPV declines slowly.**
:::

::: {style="background:#f0f4f8; border-left:4px solid #2E4057; padding:0.55em 0.8em; font-size:0.88em;"}
Implication: low-prevalence screening is most useful for **ruling conditions out** (high NPV), not ruling them in. Universal screening in MTSS Tier 1 operates in exactly this range.
:::

::: notes
The asymmetry is worth spending time on. Students who understand the algebra understand why screening programs built around sensitivity/NPV are rational for rare conditions.
:::

---

## Base rates across settings

```{r}
tribble(
  ~Setting,                      ~`Typical base rate`,               ~Implication,
  "General education classroom", "5-10% for any given condition",    "Positive universal screen often a false positive; confirmation needed",
  "MTSS Tier 2 (selected)",      "20-35%",                           "Positive screen more meaningful; still benefit from verification",
  "Outpatient referral clinic",  "25-50%",                           "PPV substantial; screening and diagnostic assessment aligned",
  "Specialty mood disorder clinic","50-70%",                         "High PPV; negative result more notable than positive",
  "Population health screening", "1-5%",                             "Even specific tests generate false positives; two-stage design required"
) |>
  kable() |>
  kable_styling(bootstrap_options=c("striped","hover"),
                full_width=TRUE, font_size=17) |>
  row_spec(0, background="#2E4057", color="white") |>
  row_spec(1, background="#FCE4D6") |>
  row_spec(3:4, background="#E8F5E9")
```

::: notes
School psych students recognize Tier 1/2 framing. Clinical students recognize the outpatient vs. specialty distinction. The red row is where most school psychology universal screening happens.
:::

# Likelihood Ratios {background-color="#2E4057"}

## Why likelihood ratios solve the prevalence problem

::: {style="font-size:0.90em; margin-bottom:0.5em;"}
PPV and NPV are setting-specific. A PPV computed in a specialty clinic cannot be applied in a school or primary care context.

**Likelihood ratios** express diagnostic weight in a form independent of prevalence, combinable with any base rate.

$$LR+ = \frac{\text{Sensitivity}}{1 - \text{Specificity}} \qquad LR- = \frac{1 - \text{Sensitivity}}{\text{Specificity}}$$

They operate on the **odds scale**. Bayes' theorem in odds form:

$$\text{Post-test odds} = \text{Pre-test odds} \times LR$$

Convert prevalence to odds, multiply by LR, convert back to probability. The ratio of hit rate to false alarm rate does not change when the population changes — **that is why LRs generalize.**
:::

::: notes
Students often treat LR multiplication as a formula to memorize. The sentence about the ratio not changing when the population changes is the conceptual key.
:::

---

## LR interpretation benchmarks

```{r}
tribble(
  ~LR,               ~`Change in probability`,   ~Interpretation,
  "LR+ > 10",        "Large (often decisive)",    "Strongly supports presence",
  "LR+ 5-10",        "Moderate",                  "Useful diagnostic shift",
  "LR+ 2-5",         "Small",                     "Minor but sometimes important",
  "LR = 1.0",        "None",                      "Test adds no information",
  "LR- 0.2-0.5",     "Small",                     "Minor reduction in probability",
  "LR- 0.1-0.2",     "Moderate",                  "Useful diagnostic shift",
  "LR- < 0.1",       "Large (often decisive)",    "Strongly supports absence"
) |>
  kable() |>
  kable_styling(bootstrap_options=c("striped","hover"),
                full_width=TRUE, font_size=20) |>
  row_spec(0, background="#2E4057", color="white") |>
  row_spec(4, background="#FFF2CC") |>
  column_spec(1, bold=TRUE)
```

::: {style="font-size:0.82em; color:#555; margin-top:0.5em;"}
PHQ-9 analog: LR+ = **`r round(phq_st$lrp,1)`** (useful), LR- = **`r round(phq_st$lrn,3)`** (good). COVID antigen: LR+ = **`r round(cov_st$lrp,1)`** (excellent), LR- = **`r round(cov_st$lrn,3)`** (modest).
:::

::: notes
The contrast between COVID LR+ and LR- is intuitive — excellent specificity gives a huge positive LR but sensitivity is only .86 so the negative LR is modest.
:::

---

## Fagan nomogram: outpatient setting (25% base rate)

```{r fig-nom-out, fig.height=5.2, fig.width=5.8, out.width="56%"}
nomogrammer(Prevalence=0.25, Sens=0.88, Spec=0.88,
            Detail=TRUE, NullLine=TRUE, LabelSize=2.4)
```

::: {style="font-size:0.90em; margin-top:0.4em;"}
**Pre-test 25%.** Positive result (blue line) updates to **~71%**. Negative result (orange line) updates to **~4%**.
Draw a line from pre-test probability through the LR, read off post-test probability on the right.
:::

::: notes
Walk through nomogram reading. The dashed horizontal at LR = 1 shows a null test that adds no information. Positive screen in outpatient context: actionable but not diagnostic on its own.
:::

---

## Fagan nomogram: community screening (5% base rate)

```{r fig-nom-comm, fig.height=5.2, fig.width=5.8, out.width="56%"}
nomogrammer(Prevalence=0.05, Sens=0.88, Spec=0.88,
            Detail=TRUE, NullLine=TRUE, LabelSize=2.4)
```

::: {style="font-size:0.90em; margin-top:0.4em;"}
**Same likelihood ratios. Pre-test 5%.** Positive result updates to only **~28%**. Negative result updates to **~1%**.
The lines are identical — what changes is where they start.
:::

::: notes
This is the single most important slide for school psychologists. A positive universal screen is not the same thing as a positive screen in a referred patient. Same test, same LRs, completely different clinical meaning.
:::

---

## Putting it together: sequential Bayesian updating

::: {style="font-size:0.85em; margin-bottom:0.5em;"}
Each assessment step uses the **previous post-test probability as the new pre-test probability**.
Likelihood ratios multiply odds — they do not add probabilities.
:::

:::: {.columns}
::: {.column width="48%"}
```{r fig-nom-seq1, fig.height=4.8, fig.width=5.2, out.width="96%"}
nomogrammer(Prevalence=0.25, Plr=7.2, Nlr=0.14,
            Detail=TRUE, NullLine=TRUE, LabelSize=2.3)
```
**Step 1:** PHQ-9 positive (LR+ = 7.2)
:::

::: {.column width="48%"}
```{r fig-nom-seq2, fig.height=4.8, fig.width=5.2, out.width="96%"}
nomogrammer(Prevalence=0.71, Plr=3.3, Nlr=0.19,
            Detail=TRUE, NullLine=TRUE, LabelSize=2.3)
```
**Step 2:** MINI positive (LR+ = 3.3)
:::
::::

::: {style="background:#f0f4f8; border-left:4px solid #2E4057; padding:0.5em 0.9em; font-size:0.82em; margin-top:0.4em;"}
**Step 1:** Pre-test odds = 0.25/0.75 = 0.333 &nbsp;×&nbsp; 7.2 = 2.40 &nbsp;→&nbsp; post-test = 2.40/3.40 = **71%**

**Step 2:** Pre-test odds = 0.71/0.29 = 2.448 &nbsp;×&nbsp; 3.3 = 8.08 &nbsp;→&nbsp; post-test = 8.08/9.08 = **89%**
:::

::: notes
Walk through each step explicitly. The key point: post-test 1 becomes the new pre-test for step 2. You are not averaging two pieces of evidence — you are compounding them. A positive screen followed by a positive structured interview produces a very high posterior probability. The assumption is that the two instruments provide independent information — if they measure the same thing in the same way, this logic breaks down.
:::

---

## What if your data disagree?

::: {style="font-size:0.88em; margin-bottom:0.5em;"}
Data don't always point the same direction. The same logic applies — the MINI result updates from wherever the PHQ left you.
:::

:::: {.columns}
::: {.column width="48%"}
```{r fig-nom-seq3, fig.height=4.8, fig.width=5.2, out.width="96%"}
nomogrammer(Prevalence=0.25, Plr=7.2, Nlr=0.14,
            Detail=TRUE, NullLine=TRUE, LabelSize=2.3)
```
**Step 1:** PHQ-9 positive (LR+ = 7.2)
:::

::: {.column width="48%"}
```{r fig-nom-seq4, fig.height=4.8, fig.width=5.2, out.width="96%"}
nomogrammer(Prevalence=0.71, Plr=3.3, Nlr=0.19,
            Detail=TRUE, NullLine=TRUE, LabelSize=2.3)
```
**Step 2:** MINI **negative** (LR− = 0.19)
:::
::::

::: {style="background:#f0f4f8; border-left:4px solid #2E4057; padding:0.5em 0.9em; font-size:0.82em; margin-top:0.4em;"}
**Step 1:** Pre-test odds = 0.333 &nbsp;×&nbsp; 7.2 = 2.40 &nbsp;→&nbsp; post-test = **71%**

**Step 2:** Pre-test odds = 0.71/0.29 = 2.448 &nbsp;×&nbsp; 0.19 = 0.465 &nbsp;→&nbsp; post-test = 0.465/1.465 = **32%**
:::

::: notes
This is the clinically important slide. A positive screen drives probability to 71% — but a negative structured interview pulls it almost all the way back to baseline (32% vs. 25% starting point). The MINI negative is not just a tie-breaker; it is meaningful evidence that substantially revises the estimate downward. The student should leave this slide understanding that disagreement between instruments is not a problem to be resolved by clinical judgment alone — it is information to be processed systematically.
:::

---

## What sequential updating tells us

::: {style="font-size:0.90em;"}
**When evidence agrees** (PHQ+ → MINI+): probability compounds upward. 25% → 71% → 89%. By the end of a thorough evaluation, you can reach high confidence.

**When evidence disagrees** (PHQ+ → MINI−): the negative finding pulls back toward baseline. 25% → 71% → 32%. The structured interview is doing real diagnostic work, not just confirming the screen.

**Three implications:**

1. Screening and confirmatory assessment are not redundant — they contribute independent information at different LR levels
2. A negative finding after a positive screen is not a failure — it is the system working correctly
3. The order matters only in terms of what probability you carry into each step — the final result is the same regardless of sequence
:::

::: {style="background:#f0f4f8; border-left:4px solid #2E4057; padding:0.5em 0.8em; font-size:0.84em; margin-top:0.5em;"}
**Assumption:** this logic requires that the two instruments provide **independent** information. If the PHQ-9 and MINI are measuring the same construct with the same items, multiplying their LRs overcounts the evidence.
:::

::: notes
The independence assumption is worth dwelling on. In practice, a PHQ-9 and a MINI are sufficiently different in format and administration that independence is defensible. But if a clinician used two similar self-report scales, the sequential update would be misleading. This connects to the broader point about assessment battery design — instruments should be selected to provide complementary, not redundant, information.
:::

# ROC Curves and AUC {background-color="#2E4057"}

## Why ROC analysis exists

::: {style="font-size:0.95em; margin-bottom:0.6em;"}
Every statistic covered so far describes the test at **one specific cutoff**. The choice of cutoff is consequential and somewhat arbitrary. Different contexts call for different cutoffs.

**ROC analysis asks:** if we evaluated this test at *every possible cutoff simultaneously*, what does its overall discriminative ability look like?

The ROC curve plots sensitivity (y) against 1 - specificity (x) at every threshold. It traces the full trade-off space. The area under the curve summarizes it in a single number that does not depend on any particular cutoff.
:::

::: {.callout-note}
Each point on the ROC curve is one row of the cutoff table. The curve *is* the table, drawn as a path through sensitivity-specificity space.
:::

::: notes
This slide lands as "why didn't we start here?" The answer: the 2x2 table and sensitivity/specificity are foundational; ROC is the generalization. The callout connecting the curve to the table they already have is the pedagogical bridge.
:::

---

## ROC curve: PHQ-9 analog

```{r fig-roc-slide, fig.height=4.8, fig.width=6.2, out.width="62%"}
roc_df <- tibble(fpr=1-roc_phq$specificities, tpr=roc_phq$sensitivities)
cut10_idx <- which.min(abs(roc_phq$thresholds - 10))
op_fpr <- 1 - roc_phq$specificities[cut10_idx]
op_tpr <- roc_phq$sensitivities[cut10_idx]

ggplot(roc_df, aes(x=fpr, y=tpr)) +
  geom_abline(slope=1, intercept=0, linetype="dashed",
              color="#aaa", linewidth=0.8) +
  geom_path(color="#2E4057", linewidth=1.5) +
  geom_point(aes(x=op_fpr, y=op_tpr),
             color="#E07B39", size=5, shape=17) +
  annotate("text", x=op_fpr+0.04, y=op_tpr-0.07,
           label=paste0("Cutoff = 10\nSe=",round(op_tpr,2),
                        ", Sp=",round(1-op_fpr,2)),
           color="#E07B39", size=3.5, hjust=0) +
  annotate("text", x=0.58, y=0.13,
           label=paste0("AUC = ",round(auc_val,3),
                        "\n95% CI: ",round(ci_val[1],3),
                        " - ",round(ci_val[3],3)),
           color="#2E4057", size=3.5, hjust=0) +
  scale_x_continuous(limits=c(0,1), breaks=seq(0,1,0.2),
                     labels=percent_format(accuracy=1)) +
  scale_y_continuous(limits=c(0,1), breaks=seq(0,1,0.2),
                     labels=percent_format(accuracy=1)) +
  labs(x="False positive rate (1 - Specificity)",
       y="True positive rate (Sensitivity)") +
  theme_minimal(base_size=13) +
  theme(panel.grid.minor=element_blank())
```

::: {style="font-size:0.88em; margin-top:0.2em;"}
AUC = **`r round(auc_val,3)`** — excellent discrimination. Orange triangle: standard cutoff of 10. Other operating points are available for contexts requiring a different balance.
:::

::: notes
Diagonal = chance (AUC = .50). Upper-left corner = perfect (AUC = 1.0). AUC = .95 means in 95% of randomly drawn depressed/non-depressed pairs, the depressed person scores higher.
:::

---

## AUC: what it means and what it does not

:::: {.columns}
::: {.column width="52%"}
**What AUC is**

The probability that a randomly selected person with the condition scores higher than a randomly selected person without it.

Equivalent to the Wilcoxon-Mann-Whitney U statistic.

Independent of prevalence and cutoff — the right metric for comparing two tests.
:::

::: {.column width="48%"}
**What AUC is not**

A statement about performance at any specific cutoff. AUC = .90 can coexist with Se = .60, Sp = .85 at the threshold actually used.

A guarantee of correct individual classification. It is a summary over all thresholds and all possible case/non-case pairs.
:::
::::

::: {style="font-size:0.82em; color:#555; margin-top:0.7em; text-align:center;"}
Benchmarks: AUC ≥ .90 excellent · .80–.89 good · .70–.79 acceptable · < .70 poor
:::

::: notes
The right column is important. Students who see "AUC = .92" in a test manual should not assume the test classifies individuals well at the cutoff they plan to use.
:::

# Cut-Score Selection {background-color="#2E4057"}

## The Youden index

::: {style="font-size:0.90em; margin-bottom:0.5em;"}
The ROC curve shows all possible operating points. The **Youden index** provides a criterion-neutral way to choose:

$$J = \text{Sensitivity} + \text{Specificity} - 1$$

Maximizing J treats false positives and false negatives as equally costly. J = 0: no better than chance. J = 1: perfect.
:::

```{r}
lapply(c(7,8,9,10,11,12,13), function(co) {
  tp_c <- sum(dat_phq$true_status=="Depressed"    & dat_phq$score>=co)
  fn_c <- sum(dat_phq$true_status=="Depressed"    & dat_phq$score<co)
  fp_c <- sum(dat_phq$true_status=="Not depressed" & dat_phq$score>=co)
  tn_c <- sum(dat_phq$true_status=="Not depressed" & dat_phq$score<co)
  st   <- compute_dx_stats(tp_c,fp_c,fn_c,tn_c,prevalence=0.25)
  tibble(Cutoff=co, Se=round(st$sens,2), Sp=round(st$spec,2),
         PPV=round(st$ppv,2), NPV=round(st$npv,2),
         J=round(st$sens+st$spec-1,3))
}) |>
  bind_rows() |>
  kable(col.names=c("Cutoff","Se","Sp","PPV","NPV","Youden J")) |>
  kable_styling(bootstrap_options=c("striped","hover"),
                full_width=FALSE, font_size=19) |>
  row_spec(0, background="#2E4057", color="white") |>
  row_spec(4, background="#D6EAF8") |>   # cutoff 10 (standard)
  row_spec(which.max(sapply(7:13, function(co) {  # Youden max row
    tp_c <- sum(dat_phq$true_status=="Depressed"    & dat_phq$score>=co)
    fn_c <- sum(dat_phq$true_status=="Depressed"    & dat_phq$score<co)
    fp_c <- sum(dat_phq$true_status=="Not depressed" & dat_phq$score>=co)
    tn_c <- sum(dat_phq$true_status=="Not depressed" & dat_phq$score<co)
    st   <- compute_dx_stats(tp_c,fp_c,fn_c,tn_c,prevalence=0.25)
    st$sens+st$spec-1
  })), background="#FFF2CC")
```

::: notes
Blue row: standard clinical cutoff (10). Yellow row: Youden-optimal. Published cutoffs often predate formal optimization — this is common in psychological assessment.
:::

---

## When Youden is not the right criterion

```{r}
tribble(
  ~Context,                       ~Priority,               ~Rationale,
  "Suicide risk screening",        "Maximize sensitivity",  "Missed case: catastrophic. False positive: manageable.",
  "MTSS Tier 1 universal",         "Balanced or favor Se",  "False negatives leave children without support",
  "Special education referral",    "Favor specificity",     "False positive triggers costly, stigmatizing evaluation",
  "Forensic PVT",                  "Fix Sp >= .90",         "False positive (genuine patient = feigning) unacceptable",
  "Community MH screening",        "Very high NPV",         "Goal is ruling out; positives require follow-up anyway"
) |>
  kable() |>
  kable_styling(bootstrap_options=c("striped","hover"),
                full_width=TRUE, font_size=18) |>
  row_spec(0, background="#2E4057", color="white") |>
  row_spec(1, background="#FCE4D6") |>
  row_spec(c(3,4), background="#FFF2CC")
```

::: notes
Red row (suicide risk) and yellow rows (special education, forensic) represent the two poles. Students should be able to articulate why the optimal threshold differs across contexts without any arithmetic.
:::

---

## The utility threshold formula

::: {style="font-size:0.90em; margin-bottom:0.5em;"}
The cost-benefit reasoning can be made explicit. The optimal probability threshold is:

$$p^* = \frac{C_{FP}}{C_{FP} + C_{FN}}$$

where $C_{FP}$ is the cost of a false positive and $C_{FN}$ is the cost of a false negative.
:::

```{r}
tribble(
  ~Scenario,            ~`Cost FP`, ~`Cost FN`, ~`Optimal threshold`,
  "Equal costs",         "1",        "1",         "p* = .50",
  "FN 10x costlier",     "1",        "10",        "p* = .09",
  "FP 10x costlier",     "10",       "1",         "p* = .91"
) |>
  kable() |>
  kable_styling(bootstrap_options=c("striped","hover"),
                full_width=FALSE, font_size=21) |>
  row_spec(0, background="#2E4057", color="white") |>
  row_spec(2, background="#FCE4D6")
```

::: {style="font-size:0.85em; color:#555; margin-top:0.5em;"}
Youden-optimal corresponds roughly to equal costs. Suicide risk screening, where missed cases may result in death, can justify p* near .05 or lower.
:::

::: notes
This is the formal underpinning of the "it depends on context" argument. The formula shows the choice is principled — it requires making the cost structure explicit, which is a clinical and ethical judgment, not a statistical one.
:::


# Calibration {background-color="#2E4057"}


## Discrimination vs. calibration

::: {style="font-size:0.92em; margin-bottom:0.6em;"}
**Discrimination** (AUC): can the test rank-order individuals correctly by condition status?

**Calibration**: when the model assigns a 50% probability to a set of individuals, does approximately 50% of them actually have the condition?

These are **independent properties**. A test can discriminate excellently and be badly miscalibrated — correctly separating cases from non-cases in rank order while systematically overestimating or underestimating absolute probabilities.
:::

::: {.callout-note}
Calibration matters most when probabilities are used as inputs to clinical decisions — exactly the nomogram workflow we just covered.
:::

::: notes
Many students have never encountered calibration as distinct from discrimination. The nomogram connection is key: if you feed a miscalibrated probability into a Bayesian update, the output will be wrong in a systematic, predictable direction.
:::

---

## Calibration plot: PHQ-9 analog

```{r fig-cal-slide, fig.height=4.2, fig.width=5.8, out.width="58%"}
mod_cal  <- glm(I(true_status=="Depressed") ~ score,
                data=dat_phq, family=binomial)
dat_phq2 <- dat_phq |>
  mutate(pred_prob=predict(mod_cal, type="response"))

cal_df <- dat_phq2 |>
  mutate(decile=ntile(pred_prob, 10)) |>
  group_by(decile) |>
  summarise(mean_pred=mean(pred_prob),
            obs_rate=mean(true_status=="Depressed"),
            .groups="drop")

# Fit a straight line through the decile points for the reference line
# (the loess S-curve is an artifact of the logistic fit on its own training data)
cal_lm <- lm(obs_rate ~ mean_pred, data=cal_df)

ggplot(cal_df, aes(x=mean_pred, y=obs_rate)) +
  geom_abline(slope=1, intercept=0, linetype="dashed",
              color="#aaa", linewidth=1.0) +
  geom_smooth(method="lm", se=FALSE,
              color="#888", linewidth=0.9, linetype="solid") +
  geom_point(size=4.5, color="#2E4057") +
  scale_x_continuous(limits=c(0,1), breaks=seq(0,1,0.2),
                     labels=percent_format(accuracy=1)) +
  scale_y_continuous(limits=c(0,1), breaks=seq(0,1,0.2),
                     labels=percent_format(accuracy=1)) +
  labs(x="Mean predicted probability (by decile)",
       y="Observed proportion depressed") +
  theme_minimal(base_size=13) +
  theme(panel.grid.minor=element_blank())
```

::: {style="font-size:0.85em; margin-top:0.3em;"}
Dashed line = perfect calibration. Points near the diagonal indicate good calibration. The solid line is a linear fit through the decile points.
:::

::: notes
Loess replaced with a linear fit to avoid the misleading S-curve artifact from fitting a logistic model on its own training data. In real validation studies, calibration plots frequently show systematic departures — overconfidence (steep S-curve) or underconfidence (shallow slope).
:::

# Common Misconceptions {background-color="#2E4057"}

## Five things that are commonly wrong

::: {style="font-size:0.89em;"}
**1. High sensitivity does not mean few false positives.**
False positives are determined by 1 - specificity (the false positive rate), not by sensitivity.

**2. High specificity does not guarantee high PPV.**
PPV depends on prevalence. At low base rates, even a 99% specific test can produce mostly false positives.

**3. AUC is not a statement about performance at any specific cutoff.**
AUC = .90 says nothing about what happens at the threshold actually used in practice.

**4. Overall accuracy is almost always misleading.**
It is dominated by the most common outcome, not by the clinically important cases.

**5. Sensitivity/specificity and PPV/NPV answer different questions.**
They are conditioned on different things. Conflating them is one of the most consequential interpretive errors in applied assessment.
:::

::: notes
Works well as a pair exercise: put up the slide, cover the explanations, ask students to articulate what is wrong with each claim before revealing it.
:::


# What This Means for Practice {background-color="#2E4057"}


## What test manuals report vs. what you need

```{r}
tribble(
  ~`Typically reported`,               ~`Rarely present`,
  "Internal consistency (alpha)",       "Sensitivity and specificity at multiple cutoffs",
  "Factor structure",                   "PPV and NPV at multiple prevalence levels",
  "Criterion correlations",             "AUC with confidence intervals",
  "Normative data",                     "Likelihood ratios",
  "Overall classification accuracy",    "Calibration evidence",
  "(sometimes) Hit rate",               "Description of the criterion standard"
) |>
  kable() |>
  kable_styling(bootstrap_options=c("striped","hover"),
                full_width=TRUE, font_size=19) |>
  row_spec(0, background="#2E4057", color="white") |>
  column_spec(1, color="#666") |>
  column_spec(2, color="#2E4057", bold=TRUE)
```

::: notes
Directly connected to the Test Review assignment. The absence of the right column is an evidentiary gap, not evidence of absence.
:::

---

## Questions to ask in a test review

::: {style="font-size:0.88em;"}
- Was a **reference standard** identified, and is it defensible?
- Were **sensitivity and specificity** reported at the primary cutoff, **with confidence intervals**?
- Were operating characteristics reported **across multiple cutoffs**?
- Was **AUC** reported with a confidence interval?
- Were **PPV and NPV** reported with the **prevalence value** used to compute them specified?
- Were **likelihood ratios** reported?
- Was **calibration** evaluated?
- Were results **stratified by demographic group**?
- Was the **study sample** appropriate for the population with whom the test will be used?
:::

::: {style="font-size:0.84em; color:#555; margin-top:0.6em; border-top:1px solid #ddd; padding-top:0.4em;"}
A statement that "the test correctly classified X% of cases" answers one of these questions — and it is the least informative one.
:::

::: notes
Directly actionable for the Test Review assignment. Students can use this as a checklist. The closing line is the chapter's thesis statement.
:::

---

## Summary

::: {style="font-size:0.86em;"}
**Sensitivity/specificity** are conditioned on true status. They describe test performance, not decision performance. They do not depend on prevalence.

**PPV/NPV** are conditioned on test result. They answer the clinical question but are setting-specific — they depend on base rate.

**Likelihood ratios** generalize across settings by operating on the odds scale. The Fagan nomogram implements Bayesian updating graphically.

**AUC** provides a cutoff-independent summary of discriminative ability. It is not a statement about any specific threshold.

**Calibration** — whether predicted probabilities match observed rates — is independent of discrimination and matters when probabilities drive decisions.

**Cut-score selection** is a values judgment about the relative costs of false positives and false negatives. Youden's J is the criterion-neutral default. The optimal threshold differs by setting, population, and purpose.

**Most test manuals report none of this.** That absence is an evidentiary gap.
:::

::: notes
Reference slide — each bullet maps to one section. The closing line is the chapter's take-home.
:::

---

## Reading

::: {style="font-size:0.88em;"}
**Required**

Youngstrom, E. A. (2014). A primer on receiver operating characteristic analysis and diagnostic efficiency statistics for pediatric psychology. *Journal of Pediatric Psychology*, *39*(2), 204–221.

VanDerHeyden, A. M. (2013). Universal screening may not be for everyone. *School Psychology Review*, *42*(4), 402–414.

**Recommended**

Streiner, D. L., & Cairney, J. (2007). What's under the ROC? *The Canadian Journal of Psychiatry*, *52*(2), 121–128.

Kroenke, K., Spitzer, R. L., & Williams, J. B. W. (2001). The PHQ-9. *Journal of General Internal Medicine*, *16*(9), 606–613.
:::

::: notes
Required readings match the syllabus. The PHQ-9 paper lets students compare what Kroenke et al. actually report against the checklist on the previous slide.
:::
